---
title: "R Notebook"
output:
  html_notebook:
    theme: united
    toc: yes
    code_folding: hide
---

The goal is to use bayesian regression to estimate codon specific elongation speeds. Gamma prior is chosen as default, since it was found to be the best prior in a previous analysis. Consensus ES is used to estimate the codon specific elongation speeds, and there is also a focus on two different CDS bins of interest, namely, 400-800 and 800-1200 bp.

# Loading files and libraries

```{r}
library(tidyverse)
library(brms)
library(cmdstanr)
library(ggpubr)
```

```{r}
codon_count <- readxl::read_excel("data_files/Codon_counts.xlsx")[c(1:6600),] %>% column_to_rownames("...1")

codon_freq <- codon_count %>% 
  mutate(across(where(is.numeric))/rowSums(across(where(is.numeric)))) %>%
  rownames_to_column("Gene")

TE <- readxl::read_excel("Unification_wo_normalisation_TE.xlsx")
```

# Pre-regression analysis

**Keeping only the 61 sense codons**

```{r}
codon_freq <- codon_freq %>% select(-c(TAG,TGA,TAA))
```

```{r}
TE <- TE %>% drop_na(Consensus_TE_speed_aa_sec) %>%
  mutate(CDS_bins_400 = cut(CDSLength,breaks=c(min(CDSLength),
                                               400,
                                               800,
                                               1200,
                                               1800,
                                               max(CDSLength)),
                            include.lowest = TRUE))
```

## bin II: 400-800 bp

bin number based on original division of genes in CDS bins

```{r}
T_II <- TE %>% 
  filter(CDS_bins_400 == "(400,800]") %>% 
  select(SystematicName,Consensus_TE_speed_aa_sec)
T_II <- merge(T_II,codon_freq, by.x = "SystematicName", by.y= "Gene")
T_II <- T_II %>% column_to_rownames("SystematicName")
```

## bin III: 800-1200 bp

bin number based on original division of genes in CDS bins

```{r}
T_III <- TE %>% 
  filter(CDS_bins_400 == "(800,1.2e+03]") %>% 
  select(SystematicName,Consensus_TE_speed_aa_sec)
T_III <- merge(T_III,codon_freq, by.x = "SystematicName", by.y= "Gene")
T_III <- T_III %>% column_to_rownames("SystematicName")
```

# Bayesian regression

Stictly positive and continuous as justification for family gamma term

Strictly positive values as well as constraints added to the output (so that it is not full of v. high or v. low values) as justification for prior gamma term

## Parameters

Change number of threads as necessary

```{r}
positive_prior <- prior(gamma(2, 1), class = "b", lb=0)
num_threads <- threading(7)
```

## bin II: 400-800 bp

```{r}
brms_model_II <- brm(bf(Consensus_TE_speed_aa_sec ~ .),
  data = T_II,
  family = Gamma(link = "log"),
  prior = positive_prior,
  backend = "cmdstanr",
  threads = num_threads,
  chains = 4,
  iter = 4000)
```

```{r}
codon_speed_II <- summary(brms_model_II)
codon_speed_II <- codon_speed_II$fixed %>% 
  select(Estimate,Est.Error,`l-95% CI`,`u-95% CI`) %>% 
  rownames_to_column("Codon") %>%
  filter(Codon != "Intercept")
```

```{r}
gghistogram(data = codon_speed_II, x = "Estimate",
            fill = "lightgrey", add = "mean",
            y = "density",
            add_density = TRUE,
            title = "Histogram of codon level speeds",
            xlab = "Codon specific elongation speed (1/s)", 
            ylab = "Number of codons",
            rug = TRUE)
```

```{r}
median(codon_speed_II$Estimate)
```

## bin III: 800-1200 bp

```{r}
brms_model_III <- brm(bf(Consensus_TE_speed_aa_sec ~ .),
  data = T_III,
  family = Gamma(link = "log"),
  prior = positive_prior,
  backend = "cmdstanr",
  threads = num_threads,
  chains = 4,
  iter = 4000)
```

```{r}
codon_speed_III <- summary(brms_model_III)
codon_speed_III <- codon_speed_III$fixed %>% 
  select(Estimate,Est.Error,`l-95% CI`,`u-95% CI`) %>% 
  rownames_to_column("Codon") %>%
  filter(Codon != "Intercept")
```

```{r}
codon_speed_III <- summary(brms_model_III)
codon_speed_III <- codon_speed_III$fixed %>% 
  select(Estimate,Est.Error,`l-95% CI`,`u-95% CI`) %>% 
  rownames_to_column("Codon") %>%
  filter(Codon != "Intercept")
```

```{r}
gghistogram(data = codon_speed_III, x = "Estimate",
            fill = "lightgrey", add = "mean",
            add_density = TRUE, #bins = 50,
            title = "Histogram of codon level speeds",
            xlab = "Codon specific elongation speed (1/s)", 
            ylab = "Number of codons",
            rug = TRUE)
```

```{r}
median(codon_speed_III$Estimate)
```

## Scatter plot bin II vs bin III

```{r}
codon_speed_comp <- merge(codon_speed_II,
                          codon_speed_III,
                          by = "Codon", 
                          suffixes = c("_II","_III"))
```

```{r}
writexl::write_xlsx(codon_speed_comp,"Codon_Speed.xlsx")
```

```{r}
ggplot(data = codon_speed_comp, aes(x=Estimate_II, y=Estimate_III)) +
  geom_point()  +
  geom_smooth(method='lm') +
  labs(x = expression(atop("Elongation speed ("*s^-1*")",paste("400-800 nt"))),
       y = expression(atop("Elongation speed ("*s^-1*")",paste("800-1200 nt")))) +
  theme(legend.position = "right",
              legend.title = element_text(size=16,hjust = 0.5),
              legend.background = element_rect(colour="black"),
              legend.text = element_text(size=16),
              axis.text.x = element_text(size = 18, angle = 0),
              axis.text.y = element_text(size = 18),
              axis.title.x = element_text(size = 18),   
              axis.title.y = element_text(size = 18), 
              panel.background = element_blank(),
              panel.grid = element_blank(),
              axis.line = element_line(colour = "black")) +
  annotate(geom = "text", 
           x = 2.5, y = 14, 
           label = "Codon-specific",
           size = 8)
```
